`State and Parameter Estimation for Spatio-Temporal Neural Fields' or `A Data Driven Framework for Patient-Specific Neural Field Modelling'

### 2010-06-17 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
changed table1 layout
changed the text in fig4b and fig4c
for merge

The sensors were spaced on a regular square 14 $\times$ 14 grid, the sensor width parameter $\sigma^2_m$ was $0.81$, which provides a width of $1.5$~mm at half the maximum amplitude of the basis function, with spacing $\Delta_y=1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical for neurophysiological recordings. 
\begin{figure}
   	\begin{center}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePy.pdf}} \\
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\end{center}
   	\caption{\textbf{a.} Example of the neural feld. \textbf{b.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{c.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
\label{fig:experimental design}
\end{figure}
\subsection{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency analysis was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency of the neural was taken to be $\nu_c \approx 0.24$~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of $2.05$~mm. This confirms that the separation of $\Delta_{y} = 1.5$~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\subfigure[][]{\label{fig:FieldFFT}\includegraphics{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics{./Graph/FFTEstimatedField.pdf}}\\
\caption{Spatial frequency analysis of the neural field. \textbf{a.} The average (over time) power in dB of the spatial frequency of the neural field. \textbf{b.} The average (over time) power in dB of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
\end{figure}

The spatial frequency of the observed field is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency was taken to be $\nu_{cy} \approx 0.19$~Hz ($-3$~dB point) \dean{the Figure~\ref{fig:SensorFunctionFreqResponse} has the cut-off as $\approx 0.2$~Hz!}. The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter of $\rho_{\phi}=1$, the minimum distance between basis functions was calculated to be $2.63$~mm to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$~mm, which is sufficient to capture the significant dynamics according to our criteria.
Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function decomposition up to a spatial frequency $\mathbf{\nu_{c\phi}} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
	\begin{center}
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\end{center}
	\caption{\dean{Note: we use $\nu_{cy}$ and $\nu_{c\phi}$ in the labels on the graphs.} Spatial frequency analysis. The dashed line shows the cut-off frequency (-3 dB point). \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 
\end{figure}

@@

Histograms of the parameter estimates obtained over the 100 realisations are shown in Figure~\ref{fig:Parameters}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given in Table~\ref{tab:Parameters estimates}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good concordance with the actual values. 
% It should be noted that when the excitatory connectivity parameters, $\hat\theta_0$ and $\hat\theta_2$, were under (or over) estimated, the inhibitory connectivity parameter, $\hat\theta_1$, tended to be over (or under) estimated. This is demonstrated in Figure~\ref{fig:ParametersRatio} where we show the distribution of the ratios of the parameter estimates. 
\begin{table}
\centering
\begin{tabular}{cccc}
	\hline\hline Parameter & Actual Value & Mean Estimate $\pm$ Std. Dev. & Bias (\%) \\
	\hline\hline
 	$\xi$ & $0.9$ & $0.925\pm 0.004$ & $2.778$ \\
	$\theta_0$ & $10$ & $10.295\pm2.879$ & $2.95$\\
	$\theta_1$ & $-8$ & $-8.571\pm1.996$ & $-7.137$\\
	$\theta_2$ & $0.5$ & $0.551\pm0.083$ & $10.2$\\
	% $\hat\theta_0/\hat\theta_1$ & $1.25$ & $?\pm0.0812$ &\\
	% $\hat\theta_0/\hat\theta_2$ & $20$ & $?\pm3.757$ & \\
\hline\hline
\end{tabular}
\caption{Actual parameter values compared to parameter estimates.}
\label{tab:Parameters estimates}
\end{table}
\begin{figure}
    \centering
    \subfigure[][]{\includegraphics{./Graph/zeta.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta0.pdf}}\\
    \subfigure[][]{\label{fig:theta1dis}\includegraphics{./Graph/theta1.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta2.pdf}}
\caption{Distribution of the parameter estimates over 100
realisations. The dotted line shows the actual value of the parameters and the solid line shows the mean of the estimated parameters. \textbf{a.} The distribution of the $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant. \textbf{b.} The distribution of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. \textbf{c.} The distribution of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. \textbf{d.} The distribution of the longer range excitatory connectivity kernel basis function parameter estimates, $\theta_2$.}
\label{fig:Parameters}
\end{figure}
\begin{figure}
    \centering
\subfigure[][]{\includegraphics{./Graph/theta0_theta1_ratio.pdf}}
\subfigure[][]{\includegraphics{./Graph/theta0_theta2_ratio.pdf}}\\
\caption{\dean{Note: need to put $\left| \right|$ around $\hat\theta_0/\hat\theta_1$ in xlabel in \textbf{a.}} Distribution of the connectivity kernel basis function parameter ratios over 100 realisations. \textbf{a.} The absolute value of the ratio of the local excitatory parameter to the surround inhibition parameter (true ratio is 1.25). \textbf{b.} The ratio of the local excitatory parameter to the longer range excitatory parameter (true ratio is 20). The dotted line shows the location of the true ratio in each case.}
\label{fig:ParametersRatio}
\end{figure}

The reconstructed connectivity kernels from the estimated parameters are illustrated in Figure~\ref{fig:KernelEstimates}, where the mean over the 100 realisations is shown as a solid line and the range indicated by the shaded region, which is  the $95\%$ confidence interval. The figure verifies that the shape of the estimated connectivity kernel is in good accordance to the true kernel. The true kernel lies within the confidence interval; however, the mean estimated kernel is more negative than the actual kernel due to the bias on the estimates of the inhibitory parameter, $\hat\theta_1$, as seen in Figure~\ref{fig:theta1dis}.
\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
% \subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Connectivity kernel estimate. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the $95~\%$ confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}

The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over time) of the root mean square error (MRMSE) (over space). The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. The error is consistently around 0.5 over time, which approximately \dean{??}~\% of the voltages, which range over \dean{(??,??)}. The error in the reconstructed field is mostly due to the loss high spatial frequency information due to the basis function decomposition; this is illustrated in Figure~\ref{fig:FieldEstimate}, which shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Fig 10b that the estimated field contains the basic structure of the true field (fig10a), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). } 
\label{fig:RMSE}
   \end{figure}

>>>>

The sensors were spaced on a regular square $14 \times 14$ grid, the sensor width parameter $\sigma^2_m$ was set to $0.81$, which provides a width of $1.5$~mm at half the maximum amplitude of the basis function, with spacing $\Delta_y=1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical for neurophysiological recordings. 
\begin{figure}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\caption{\textbf{a.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
\label{fig:experimental design}
\subsection{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency analysis was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency of the neural was taken to be $\nu_c \approx 0.24$~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of $2.05$~mm. This confirms that the separation of $\Delta_{y} = 1.5$~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\subfigure[][]{\label{fig:FieldFFT}\includegraphics{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics{./Graph/FFTEstimatedField.pdf}}\\
\caption{Spatial frequency analysis of the neural field. \textbf{a.} The average (over time) power in dB of the spatial frequency of the neural field. \textbf{b.} The average (over time) power in dB of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
The spatial frequency of the observed field is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency was taken to be $\nu_{cy} \approx 0.19$~Hz ($-3$~dB point) \dean{the Figure~\ref{fig:SensorFunctionFreqResponse} has the cut-off as $\approx 0.2$~Hz!}. The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter of $\rho_{\phi}=1$, the minimum distance between basis functions was calculated to be $2.63$~mm to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$~mm, which is sufficient to capture the significant dynamics according to our criteria.
Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function decomposition up to a spatial frequency $\mathbf{\nu_{c\phi}} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\caption{\dean{Note: we use $\nu_{cy}$ and $\nu_{c\phi}$ in the labels on the graphs.} Spatial frequency analysis. The dashed line shows the cut-off frequency (-3 dB point). \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the one-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a one-dimensional model sensor. \textbf{c.} The spatial frequency response of a one-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 

@@

Histograms of the parameter estimates obtained over the 100 realisations are shown in Figure~\ref{fig:Parameters}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given in Table~\ref{tab:Parameters estimates}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good concordance with the actual values. 
% It should be noted that when the excitatory connectivity parameters, $\hat\theta_0$ and $\hat\theta_2$, were under (or over) estimated, the inhibitory connectivity parameter, $\hat\theta_1$, tended to be over (or under) estimated. This is demonstrated in Figure~\ref{fig:ParametersRatio} where we show the distribution of the ratios of the parameter estimates. 
\begin{table}
\centering
\begin{tabular}{cccc}
	\hline\hline Parameter & Actual Value & Mean Estimate $\pm$ Std. Dev. & Bias (\%) \\
	\hline\hline
 	$\xi$ & $0.9$ & $0.925\pm 0.004$ & $2.778$ \\
	$\theta_0$ & $10$ & $10.295\pm2.879$ & $2.95$\\
	$\theta_1$ & $-8$ & $-8.571\pm1.996$ & $-7.137$\\
	$\theta_2$ & $0.5$ & $0.551\pm0.083$ & $10.2$\\
	% $\hat\theta_0/\hat\theta_1$ & $1.25$ & $?\pm0.0812$ &\\
	% $\hat\theta_0/\hat\theta_2$ & $20$ & $?\pm3.757$ & \\
\hline\hline
\end{tabular}
\caption{Actual parameter values compared to parameter estimates.}
\label{tab:Parameters estimates}
\end{table}
\begin{figure}
    \centering
    \subfigure[][]{\includegraphics{./Graph/zeta.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta0.pdf}}\\
    \subfigure[][]{\label{fig:theta1dis}\includegraphics{./Graph/theta1.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta2.pdf}}
\caption{Distribution of the parameter estimates over 100
realisations. The dotted line shows the actual value of the parameters and the solid line shows the mean of the estimated parameters. \textbf{a.} The distribution of the $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant. \textbf{b.} The distribution of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. \textbf{c.} The distribution of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. \textbf{d.} The distribution of the longer range excitatory connectivity kernel basis function parameter estimates, $\theta_2$.}
\label{fig:Parameters}
\end{figure}
\begin{figure}
    \centering
\subfigure[][]{\includegraphics{./Graph/theta0_theta1_ratio.pdf}}
\subfigure[][]{\includegraphics{./Graph/theta0_theta2_ratio.pdf}}\\
\caption{\dean{Note: need to put $\left| \right|$ around $\hat\theta_0/\hat\theta_1$ in xlabel in \textbf{a.}} Distribution of the connectivity kernel basis function parameter ratios over 100 realisations. \textbf{a.} The absolute value of the ratio of the local excitatory parameter to the surround inhibition parameter (true ratio is 1.25). \textbf{b.} The ratio of the local excitatory parameter to the longer range excitatory parameter (true ratio is 20). The dotted line shows the location of the true ratio in each case.}
\label{fig:ParametersRatio}
\end{figure}
The reconstructed connectivity kernels from the estimated parameters are illustrated in Figure~\ref{fig:KernelEstimates}, where the mean over the 100 realisations is shown as a solid line and the range indicated by the shaded region, which is  the $95\%$ confidence interval. The figure verifies that the shape of the estimated connectivity kernel is in good accordance to the true kernel. The true kernel lies within the confidence interval; however, the mean estimated kernel is more negative than the actual kernel due to the small bias on the estimates of the inhibitory parameter, $\hat\theta_1$, as seen in Figure~\ref{fig:theta1dis}.

\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
% \subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Connectivity kernel estimate. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the $95~\%$ confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}
The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over time) of the root mean square error (MRMSE) (over space). The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. The error is consistently around 0.5 over time, which approximately \dean{??}~\% of the voltages, which range over \dean{(??,??)}. The error in the reconstructed field is mostly due to the loss high spatial frequency information due to the basis function decomposition; this is illustrated in Figure~\ref{fig:FieldEstimate}, which shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Fig 10b that the estimated field contains the basic structure of the true field (fig10a), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). } 
\label{fig:RMSE}
   \end{figure}
